---
title: "01-intro.Rmd"
author: "Jenny Bryan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: TRUE
---

<blockquote class="twitter-tweet" lang="en"><p>만약 정보생물학(Bioinformatics)을 배우는 생물학자에게 한마디 말을 남긴다면,
&quot;코드는 사람을 위해 작성하고, 데이터는 컴퓨터를 위해 작성하라.&quot;가 될 것이다. <br>
If I had one thing to tell biologists learning bioinformatics, it would be write code for humans, write data for computers&quot;.</p>&mdash; Vince Buffalo (@vsbuffalo) <a href="https://twitter.com/vsbuffalo/statuses/358699162679787521">2013년, 7월 20일</a></blockquote>


"코드는 사람을 위해 작성한다"의 중요한 측면은
데이터를 __깔끔__하게 만드는 것이다. (White et al 와 Wickham 참조) __깔끔__한 데이터에 대한 주요 기능(key feature)에 대한 합의가 이뤄지고 있다.


  * 각 칼럼(column, 열)이 변수다.
  * 각 행(row)은 관측점이다.

예를 들어, 그림을 만드는데 고생을 하고 있다면, 잠시 멈추고 데이터가 __깔끔__한지에 대해서 심각히 생각하기 바란다. 깔끔하지 못한 것이 자료분석과 시각화에 있어서 종종 간과되는 고통의 원인이다.


## 반지의 제왕 예제 

[반지의 제왕 3부작에서 나온 데이터](https://github.com/jennybc/lotr)에서 생성한 깔끔하지 못한 데이터의 확실한 예제가 제시된다.

```{r make-tidy-data-untidy, include = FALSE}
library(knitr)
library(dplyr)
library(tidyr)
library(xtable)

## stores path to data directory into object data_dir
source("00-set-data-dir.R")

if(!file.exists(file.path(data_dir, "lotr_clean.tsv"))) {
  download.file(paste0("https://raw.githubusercontent.com/jennybc/",
                       "lotr/master/lotr_clean.tsv"), 
                destfile = file.path(data_dir, "lotr_clean.tsv"),
                method = "curl")
}
lotr_dat <- tbl_df(read.delim(file.path(data_dir, "lotr_clean.tsv"),
                              stringsAsFactor = FALSE))
females <- c("Galadriel", "Arwen", "Lobelia Sackville-Baggins", "Rosie",
             "Mrs. Bracegirdle", "Eowyn", "Freda", "Rohan Maiden")
lotr_dat <-
  mutate(lotr_dat,
         Film = factor(Film, levels = c("The Fellowship Of The Ring",
                                        "The Two Towers",
                                        "The Return Of The King")),
         Gender = factor(ifelse(Character %in% females, "Female", "Male")),
         Race = factor(Race))
(lotr_tidy <- lotr_dat %>%
   filter(Race %in% c("Elf", "Hobbit", "Man")) %>%
   droplevels %>%
   group_by(Film, Gender, Race) %>%
   summarize(Words = sum(Words)))
(all_combns <- with(lotr_tidy,
                    expand.grid(Film = levels(Film), Race = levels(Race),
                                Gender = levels(Gender))))
lotr_tidy <- left_join(all_combns, lotr_tidy)
## 2014-11-04 working around a dplyr join bug re factor levels
lotr_tidy <- lotr_tidy %>%
  mutate(Film = factor(Film, levels = c("The Fellowship Of The Ring",
                                        "The Two Towers",
                                        "The Return Of The King")))
lotr_tidy[is.na(lotr_tidy)] <- 0
lotr_tidy
lotr_tidy <- lotr_tidy %>% arrange(Film, Race, Gender)
```

```{r echo = FALSE}
make_untidy_table <- function(film) {
  tmp <- lotr_tidy %>%
    filter(Film == film) %>%
    select(-Film)
  with(tmp,
       data.frame(Race = Race[Gender == "Female"],
                  Female = Words[Gender == "Female"],
                  Male = Words[Gender == "Male"]))
  }
```
<table border = 1>
<tr>
<td>
```{r results='asis', echo=FALSE}
print(xtable(make_untidy_table("The Fellowship Of The Ring"),
             digits = 0, caption = "The Fellowship Of The Ring"),
      caption.placement = "top", include.rownames = FALSE, type='html')
```
</td>
<td>
```{r results='asis', echo=FALSE}
print(xtable(make_untidy_table("The Two Towers"),
             digits = 0, caption = "The Two Towers"),
      caption.placement = "top", include.rownames = FALSE, type='html')
```
</td>
<td>
```{r results='asis', echo=FALSE}
print(xtable(make_untidy_table("The Return Of The King"),
             digits = 0, caption = "The Return Of The King"),
      caption.placement = "top", include.rownames = FALSE, type='html')
```
</td>
</tr>
</table>

각 영화마다 테이블이 한개 있다.
각 테이블에 등장인물 종족과 성별로 전체 대사 단어 숫자가 나타나 있다.

상기 테이블 세개를 엑셀 워크북(workbook)에 별도 워크쉬트(worksheet)로 상상할 수 있다. 혹은,
원천 데이터를 담고 있는 워크쉬트 귀퉁이 자리잡고 있을 수도 있다. 혹은 워드 프로세서나 웹페이지의 테이블로 생각할 수도 있다.  

상기 데이터는 *사람 눈알(human eyeballs)* 소비를 위한 형식으로 준비되었다(Murrell의 말을 바꾸어 표현). 예컨데, *사람*이 두개의 탑에서 여성요정이 말한 단어 숫자를 찾기에 적합한 형식이다. 하지만, 실제로 이 형식에서 *컴퓨터*가 그런 갯수 정보를 추출하기는 매우 어렵다. 더 중요하게는 갯수를 계산하거나 그래프를 그리는 것은 정말 어렵다.

## 연습과제

상기 테이블을 살펴보고 다음 질문에 답하라.


  * 남자 호빗이 말한 총 단어 숫자는 얼마인가?
  * 특정 `종족(Race)`가 영화를 지배하고 있는가?   지배 `종족(Race)`이 영화별로 차이가 있는가?

만약 더 많은 영화가 있다면, 혹은 만약 모든 `종족(Races)`(예를 들어, 난장이, 오크 등)을 포함한 갱신 데이터를 제공한다면, 여러분의 접근법은 확장성(Scale)이 얼마나 있을까요? 

## 깔끔한 반지의 제왕 데이터

다음에 동일한 데이터가 깔끔한 형식으로 표현되면 어떤 모양일지가 나타나 있다.

```{r echo = FALSE, results = 'asis'}
print(xtable(lotr_tidy, digits = 0), include.rownames = FALSE, type='html')
```
일반적으로 깔끔한 데이터는 좀더 좁고, 키가 크다.
페이지에 멋지게 맞지는 않는다. 
특정 요소는 반복이 심하다. 예를 들어, `Hobbit`.
본증적으로 __깔끔한__ 데이터를 효율적이지 못하고 때로는 추하다고 생각해서 종종 배척한다.
하지만, 데이터를 가지고 텍스트 표현하는 최종 제품을 만들지 않거나 만들 때까지는 데이터를 꼭 맞는 간결한 형식으로 보려는 갈망은 무시해라.


## 깔끔한 데이터 효익

With the data in tidy form, it's natural to *get a computer* to do further summarization or to make a figure. This assumes you're using language that is "data-aware", which R certainly is. Let's answer the questions posed above.

#### What's the total number of words spoken by male hobbits?

```{r}
aggregate(Words ~ Race * Gender, data = lotr_tidy, FUN = sum)
```

Now it takes just one line of code to compute the word total for both genders of all `Races` across all `Films`. The total number of words spoken by male hobbits is `r lotr_tidy %>% filter(Race == 'Hobbit' & Gender == 'Male') %>% summarize(sum(Words))`. It was important here to have all word counts in a single variable, within a data frame that also included variables for `Race` and `Gender`.

#### Does a certain `Race` dominate a movie? Does the dominant `Race` differ across the movies?

First, we sum across `Gender`, to obtain word counts for the different races by movie.

```{r echo = FALSE, include = FALSE}
(tmp <- lotr_tidy %>%
   group_by(Film, Gender) %>%
   summarize(Words = sum(Words)))
```

```{r}
(by_race_film <- aggregate(Words ~ Race * Film, data = lotr_tidy, FUN = sum))
```
We can stare hard at those numbers to answer the question. But even nicer is to depict the word counts we just computed in a barchart. 

```{r barchart-lotr-words-by-film-race}
library(ggplot2)
p <- ggplot(by_race_film, aes(x = Film, y = Words, fill = Race))
p + geom_bar(stat = "identity", position = "dodge") +
  coord_flip() + guides(fill = guide_legend(reverse=TRUE))
```

`Hobbits` are featured heavily in The Fellowhip of the Ring, where as `Men` had a lot more screen time in The Two Towers. They were equally prominent in the last movie, The Return of the King.

Again, it was important to have all the data in a single data frame, all word counts in a single variable, and associated variables for `Film` and `Race`.

## Take home message

Having the data in __tidy__ form was a key enabler for our data aggregations and visualization.

Tidy data is integral to efficient data analysis and visualization.

If you're skeptical about any of the above claims, it would be interesting to get the requested word counts, the barchart, or the insight gained from the chart *without* tidying or plotting the data. And imagine redoing all of that on the full dataset, which includes 3 more `Races`, e.g. `Dwarves`.

### Where to next?

In [the next lesson](02-tidy.md), we'll show how to tidy this data.

Our summing over `Gender` to get word counts for `Film * Race` was an example of __data aggregation__. The base function `aggregate()` does simple aggregation. For more flexibility, check out the packages `plyr` and `dplyr`. *point to other lessons when/if they exist?*

The figure was made with `ggplot2`, a popular package that implements the Grammar of Graphics in R.

### Resources

  * [Bad Data Handbook](http://shop.oreilly.com/product/0636920024422.do) by By Q. Ethan McCallum, published by O'Reilly.
    - Chapter 3: Data Intended for Human Consumption, Not Machine Consumption by Paul Murrell.
  * Nine simple ways to make it easier to (re)use your data by EP White, E Baldridge, ZT Brym, KJ Locey, DJ McGlinn, SR Supp. *Ideas in Ecology and Evolution* 6(2): 1–10, 2013. doi:10.4033/iee.2013.6b.6.f <http://library.queensu.ca/ojs/index.php/IEE/article/view/4608>
    - See the section "Use standard table formats"
  * Tidy data by Hadley Wickham. Journal of Statistical Software. Vol. 59, Issue 10, Sep 2014. <http://www.jstatsoft.org/v59/i10>
    - [`tidyr`](https://github.com/hadley/tidyr), an R package to tidy data.
    - R packages by the same author that do heavier lifting in the data reshaping and aggregation department include [`reshape2`](https://github.com/hadley/reshape), [`plyr`](https://github.com/hadley/plyr) and [`dplyr`](https://github.com/hadley/dplyr).
    
```{r include = FALSE}
## leaves files behind for lesson on how to tidy
for(film in c("The Fellowship Of The Ring", "The Two Towers",
              "The Return Of The King")) {
  tmp <- data.frame(Film = film, make_untidy_table(film))
  fname <- file.path(data_dir, paste0(gsub(" ", "_", film), ".csv"))
  write.table(tmp, fname, quote = FALSE, sep = ",", row.names = FALSE)
  }

## leaves files behind for exercises re: how to tidy
for(gender in c("Female", "Male")) {
  tmp <- lotr_tidy %>% filter(Gender == gender)
  tmp <- with(tmp,
              data.frame(Gender = gender,
                         Film = Film[Race == "Elf"],
                         Elf = Words[Race == "Elf"],
                         Hobbit = Words[Race == "Hobbit"],
                         Man = Words[Race == "Man"]))
  fname <- file.path(data_dir, paste0(gender, ".csv"))
  write.table(tmp, fname, quote = FALSE, sep = ",", row.names = FALSE)
  }
```

